---
title: "Color data frame output in R terminal"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Color data frame output in R terminal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

```{r echo=FALSE,include=FALSE,eval=TRUE}
options(crayon.enabled = TRUE)
options(crayon.colors = 256)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
    '</code></pre>'
  )
})

## this is an ugly, ugly hack, but otherwise crayon does not listen
num_colors <- function(forget=TRUE) 256
library(crayon)
assignInNamespace("num_colors", num_colors, pos="package:crayon")
```


```{r setup,echo=FALSE,results="hide",include=FALSE}
library(colorDF)
library(tidyverse)
library(data.table)
```

# Quick start


```{r}
data(starwars)
sw <- starwars[, c(1:3, 7:8)]
sw %>% colorDF
colorDF(sw) %>% summary
```

# Colorful data frames

Your average terminal in which you run R is capable of displaying colors,
styles and unicode characters. Wouldn't it be nice to add some color to the
data frames you are displaying? For example, that factors are shown in a
distinct color (no confusing of strings and factors any more!) or that
significant p-values are colored in red?

This was my motivation when writing this tiny package. Of course, changing
default method for printing data frames is nothing a package is allowed to
do (but read on!). However, this package defines everything you need to get
dynamic, colorful output when viewing data frames. There are two things
about colorDF which are important:

 1. colorDF *never* modifies the behavior of the data frame like object or
    its contents (i.e. it does not redefine methods like `[<-`, removes row names etc.). 
    The only two things that change are (i) the default print method
    (visualization), and (ii) the ".style" and ".coltp" attributes of the
    object, and that only if you really change the class of the object,
    which is often unnecessary.
 2. Any data frame like object can be used in colorDF, and you don't need
    to modify these objects to use the colorful visualizations.

Yes, you can color *any* object that can be cast into a data frame
with this or related functions! For example, you can apply it to both
tibbles and data.table objects:

```{r,results="hide"}
## works with standard data.frames
colorDF(mtcars)

## works with tidyverse tibbles
mtcars %>% as_tibble %>% colorDF

## works with data.table
colorDF(data.table(mtcars))
```

The output of these three commands is identical:

```{r echo=FALSE}
colorDF(mtcars)
```

# Column types

Column types are mostly like classes, but colorDF introduces some
additional distinctions, specifically "identifier" (such that character
columns which contain identifiers can be shown with a particular, distinct
style) and "pval", to show significant p-values in a different color (and
use `format.pval()` for formatting). Column types are stored in the
`.coltp` attribute of the colorDF object.

colorDF tries to guess how each column should be displayed. First it checks
whether any column types have been assigned explicitely using the
`col_type<-` function and stored in the `.coltp` attribute of the object.
Next, it looks up whether it can guess the contents of the column by
looking at the column name (ID, p-value). Finally, it determines the class
of the column (character, integer, numeric, logical, factor).  

To assign a particular column type, you need first to turn a data frame
colorful and then modify the column type:

```{r}
sw <- sw %>% as.colorDF
col_type(sw, "name") <- "identifier"
col_type(sw, "gender") <- "factor"
sw$probability <- runif(nrow(sw), 0, 0.1)
col_type(sw, "probability") <- "pval"
sw
```

Note that changing the column type **does not** change the class of the
column in the data frame! colorDF never touches the data frame contents,
the only operations concern the ".style" and ".coltp" attributes. So while
you may set a column type to "character" instead of "factor", even though
it will be looking like a character type on the terminal output, the
column class will still be a factor.

# Styles and Themes

I am a bit confused when it comes to distinguishing the two. Themes are
basically internally predefined styles. Styles are simply lists that hold
information how different columns, column and row headers, separators
between the columns and highlighted rows are displayed. 

Themes can be set using the `options(colorDF_theme="<theme name>")` command
or by directly specifying the option in a call to `colorDF`:

```{r}
colorDF(sw, theme="bw")
```

Here is an overview of the themes. Some of them are intended for dark
background and will not look great below:

```{r}
colorDF_themes_show()
```

You can add your own themes using `add_colorDF_theme()` (see the example
section on the help page).


# Column styles

Styles of a colorDF object can be directly manipulated using `df_style`:

```{r}
mtcars.c <- colorDF(mtcars)
df_style(mtcars.c, "sep") <- "; "
```

If interested, read the help file for `df_style()`.

# Utilities

colorDF comes with two utility functions. Firstly, it defines a summary
method for colorful data frames which can also be used for any other data
frame like object and which I find much more useful than the regular
summary:

```{r}
starwars %>% as.colorDF %>% summary
```

There is a directly visible (exported) version of the colorful summary
called `summary_colorDF`:

```{r eval=FALSE}
starwars %>% summary_colorDF
```

The `highlight()` function allows to mark selected rows from the table:

```{r, eval=TRUE}
foo <- starwars %>% select(name, species, homeworld) %>% 
  highlight(.$homeworld == "Tatooine")
```

(Unfortunately, the HTML representation of the ANSI terminal doesn't show
that one correctly).


# Setting up colorDF as the default data frame print method

You can use colorDF as the default method for displaying data frames and
similar objects. For this, you need to use the `colorDF:::print.colorDF`
function:

```{r eval=FALSE}
## for regular data frames
print.data.frame <- colorDF:::print.colorDF

## for tidyverse tibbles
print.tbl        <- colorDF:::print.colorDF

## for data.tables
print.data.table <- colorDF:::print.colorDF
```

This will not replace or modify the original functions from data.table or
tibble packages, but merely mask these. And from now on, every data frame
like object will be shown in color, but otherwise, its behavior will not
change.

Should you want to go back to the original print functions, just remove
these new functions:

```{r eval=FALSE}
rm(print.data.frame, print.tbl, print.data.table)
```

# Global options

There are two global options which override whatever has been defined in
a particular theme / style.

 * `colorDF_tibble_style`: if not NULL, tibble conventions will be used
   (i.e., show only columns which fit on the screen and create a subheader
   with column classes)
 * `colorDF_sep`: column separator string 
 * `colorDF_n`: how many rows to show on the screen

For example, 

```{r problemchild, eval=TRUE,results="markdown"}
options(colorDF_tibble_style="yes please")
options(colorDF_sep= " ")
options(colorDF_n=5)
colorDF(starwars)
```

# Rmarkdown

The package is intended to be used in terminal. However, as you see above,
it is possible to get the colored tables also in an rmarkdown document. For
this, include the following chunk at the beginning of your document:


````markdown
`r ''````{r echo=FALSE}
options(crayon.enabled = TRUE)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = htmltools::htmlEscape(x), warn = FALSE),
    '</code></pre>'
  )
})
```
````

# Issues

Currently, colorDF relies on the `crayon` library to generate the ANSI escape
codes. Unfortunately, crayon is peculiar about trying to guess the terminal
type. Without going into details, there are situations in which it is not
possible to force crayon into using 256 colors, even if you know that it
really works. One such example is this vignette: if an rmarkdown file is
built from command line, crayon will ignore any setting of the
`crayon.colors` option and use only the base colors.

The reason that the colors in this vignette appear correct is that I used
an ugly hack to substitute the `crayon::num_colors()` function by a
function that always returns 256:

````markdown
`r ''````{r echo=FALSE}
num_colors <- function(forget=TRUE) 256
library(crayon)
assignInNamespace("num_colors", num_colors, pos="package:crayon")
```
````



