---
title: "Color data frame output in R terminal"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
vignette: >
  %\VignetteIndexEntry{Color data frame output in R terminal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ""
)
```

```{r setup,echo=FALSE,results="hide",include=FALSE}
library(colorDF)
library(tidyverse)
library(data.table)
```

```{r echo=FALSE,include=FALSE}
options(crayon.enabled = TRUE)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = x, warn = FALSE),
    '</code></pre>'
  )
})
```

# Quick start

(Please note that the HTML colors below are only approximate rendering of how
the output will look on a terminal).


```{r}
data(starwars)
sw <- starwars[, c(1:4, 7:8)]
sw %>% colorDF
as.colorDF(sw) %>% summary
```

# Colorful data frames

Your average terminal in which you run R is capable of displaying colors,
styles and unicode characters. Wouldn't it be nice to add some color to the
data frames you are displaying? For example, that factors are shown in a
distinct color (no confusing of strings and factors any more!) or that
significant p-values are colored in red?

This was my motivation when writing this tiny package. Of course, changing
default method for printing data frames is nothing a package is allowed to
do (but read on!). However, this package defines everything you need to get
dynamic, colorful output when viewing data frames. There are two things
about colorDF which are important:

 1. colorDF *never* modifies the behavior of the data frame like object or
    its contents. The only two things that change are (i) the default print method
    (visualization), and (ii) the ".style" attribute of the object, and
    that only if you really change the class of the object, which is
    usually unnecessary.
 2. Any data frame like object can be used in colorDF, and you don't need
    to modify these objects to use the colorful visualizations.

Yes, you can color *any* object that can be cast into a data frame
with this or related functions! For example, you can apply it to both
tibbles and data.table objects:

```{r,results="hide"}
## works with standard data.frames
colorDF(mtcars)

## works with tidyverse tibbles
mtcars %>% as_tibble %>% colorDF

## works with data.table
colorDF(data.table(mtcars))
```

The output of these three commands is identical:

```{r echo=FALSE}
colorDF(mtcars)
```

# Column types

Column types are mostly like classes, but colorDF introduces some
additional distinctions, specifically "identifier" (such that character
columns which contain identifiers can be shown with a particular, distinct
style) and "pval", to show significant p-values in a different color (and
use `format.pval()` for formatting).

colorDF tries to guess how each column should be displayed. First of all,
it determines the class of the column (character, integer, numeric,
logical, factor). Next, it looks up whether it can guess the contents of
the column by looking at the column name (ID, p-value). Finally, it checks
whether any columns have been implicitely been assigned a column type.

To assign a particular column type, you need first to turn a data frame
colorful and then modify the column type:

```{r}
sw <- sw %>% as.colorDF
col_type(sw, "name") <- "identifier"
col_type(sw, "gender") <- "factor"
sw$probability <- runif(nrow(sw), 0, 0.1)
col_type(sw, "probability") <- "pval"
sw
```

Note that changing the column type **does not** change the class in the
data frame! colorDF never touches the data frame object itself, the only
operations concern the ".style" attribute. So while you may set a column
type to "character" instead of "factor", it will be looking like a
character type on the terminal output, but the column class will still be a
factor.

# Styles and Themes

I am a bit confused when it comes to distinguishing the two. Themes are
basically internally predefined styles. Styles are simply lists that hold
information how different columns, column and row headers, separators
between the columns and highlighted rows are displayed. 

Themes can be set using the `options(colorDF_theme="<theme name>")` command
or by directly specifying the option in a call to `colorDF`:

```{r}
colorDF(sw, theme="bw")
```

Here is an overview of the themes. Some of them are intended for dark
background and will not look great below:

```{r}
colorDF_themes_show()
```

You can add your own themes using `add_colorDF_theme()` (see the example
section on the help page).


# Column styles

Styles of a colorDF object can be directly manipulated using `df_style`:

```{r}
mtcars.c <- colorDF(mtcars)
df_style(mtcars.c, "sep") <- "; "
```

If interested, read the help file for `df_style()`.

# Utilities

colorDF comes with two utility functions. Firstly, it defines a summary
method for colorful data frames which can also be used for any other data
frame like object and which I find much more useful than the regular
summary:

```{r}
starwars %>% as.colorDF %>% summary
```

There is a directly visible (exported) version of the colorful summary
called `summaryColorDF`:

```{r eval=FALSE}
starwars %>% summaryColorDF
```

The `highlight()` function allows to mark selected rows from the table:

```{r, eval=FALSE}
foo <- starwars %>% select(name, species, homeworld) %>% 
  highlight(.$homeworld == "Tatooine")
```

(Unfortunately, the HTML representation of the ANSI terminal doesn't show
that one correctly).


# Rmarkdown

The package is intended to be used in terminal. However, as you see above,
it is possible to get the colored tables also in an rmarkdown document. For
this, include the following chunk at the beginning of your document:


````markdown
`r ''````{r echo=FALSE,include=FALSE}
options(crayon.enabled = TRUE)
knitr::knit_hooks$set(output = function(x, options){
  paste0(
    '<pre class="r-output"><code>',
    fansi::sgr_to_html(x = x, warn = FALSE),
    '</code></pre>'
  )
})
```
````

# Issues

The main problem is that it relies on the `crayon` package, which
unfortunately is sometimes wrong about whether the terminal has 255 colors
or 16 (at least in the version on CRAN). Case in point, the colors do not
render correctly in rstudio, despite the fact that the rstudio terminal is
of course capable of showing all ANSI colors.

